Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-04 paper/main.tex
Updates

\subsubsection{Both $ X $ and $ Y $ are continuous}
When both $ X $ and $ Y $ are continuous variables, Pearson's correlation
coefficient is typically used.  In the case when $ \bm{Z} = \emptyset $, the
correlation coefficient is:

$$ \rho_{X, Y} = \frac{cov(X, Y)}{\sigma_X \sigma_Y} $$

When $ \bm{Z} \neq \emptyset $, partial correlation coefficient can be used. 
One of the ways to estimate it is to use two linear regression models:

@@

As there are no generalized measures of association for mixed data, in this
section, we first introduce a novel measure of partial association for mixed
data by extending the concept of partial correlations \citep{Whittaker2009}
\todo{Verify this citation} to handle mixed data types. To compute partial
correlations, we have two key steps: 1) a residualization method, and 2) a
measure of associations between the residuals. When both $ X $ and $ Y $ are
continuous variables, residuals are typically computed using a regression-based
method, and Pearson's correlation coefficient is used to measure the
association between these residuals. To extend this approach to mixed data, we
combine a mixed data residualization method \citep{Ankan2023} with a
multivariate measure of association based on canonical correlations, Pillai's
Trace \citep{Pillai1955}.

Given $ n $ samples $ D = (x, y, \bm{z}) $ of variables $ X $, $ Y $, and $
\bm{Z} $, we aim to estimate the conditional association $ \rho(X, Y; \bm{Z}) $
between $ X $ and $ Y $ conditioned on $ \bm{Z} $. We use the residualization
approach from \citet{Ankan2023} to compute the residuals. For each of the
variable $ X $ and $ Y $, we compute the residual $ R_X $ and $R_Y$
respectively as follows depending on their type.

@@

We repeat the above step for the variable $ Y $ and compute its residual $ R_Y
$. Here, we have the option to choose the estimators based on the properties of
the data. Non-parametric ensemble estimators such as Random Forest and XGBoost
can handle any combination of data types while performing reasonably well. For
linear datasets, we can also choose variants of linear regression depending on
the type of data.

@@

Many measures of association based on canonical correlation have been used in
multivariate statistics, each with slightly different properties.
\begin{itemize}
	\item Wilks' Lambda:
		$$ \Lambda = \prod_{i}^{\min(p, q)} (1 - \phi_i^2) $$
	\item Roy's Largest Root: 
		$$ \theta = \max_i(\phi_i^2) $$
	\item Pillai's Trace: 
		$$ \tau = \sum_{i=1}^{\min(p, q)} \phi_i^2 $$
\end{itemize}

>>>>

\subsubsection{Both $ X $ and $ Y $ are continuous}
When both $ X $ and $ Y $ are continuous variables, Pearson's correlation
coefficient is typically used. In the case when $ \bm{Z} = \emptyset $, the
correlation coefficient is given as:

\begin{equation}
\rho_{X, Y} = \frac{\mathrm{cov}(X, Y)}{\sigma_X \sigma_Y}
\end{equation}

where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively. When $ \bm{Z} \neq \emptyset $, partial correlation coefficient
can be used. One of the ways to estimate it is to use two linear regression
models:

@@

As we saw in the previous section, depending on the type of variable, different
conditional measures of association are used. However, no generalized measure
of association exists for mixed data. In this section, we present a novel
measure of partial association for mixed data. This measure of association is
analogous to partial correlation coefficient where a residualization method is
combined with Pearson correlation coefficient. Here, we use a mixed data
residualization method \citep{Ankan2023} and combine it with a multivariate
measure of association based on canonical correlations called Pillai's Trace
{Pillai1955}.

% As there are no generalized measures of association for mixed data, in this
% section, we first introduce a novel measure of partial association for mixed
% data by extending the concept of partial correlations \citep{Whittaker2009}
% \todo{Verify this citation} to handle mixed data types. To compute partial
% correlations, we have two key steps: 1) a residualization method, and 2) a
% measure of associations between the residuals. When both $ X $ and $ Y $ are
% continuous variables, residuals are typically computed using a regression-based
% method, and Pearson's correlation coefficient is used to measure the
% association between these residuals. To extend this approach to mixed data, we
% combine a mixed data residualization method \citep{Ankan2023} with a
% multivariate measure of association based on canonical correlations, Pillai's
% Trace \citep{Pillai1955}.

Given $ n $ samples $ D = (x, y, \bm{z}) $ of variables $ X $, $ Y $, and $
\bm{Z} $, we aim to estimate the conditional association $ \rho(X, Y; \bm{Z}) $
between $ X $ and $ Y $ conditioned on $ \bm{Z} $. We first use the
residualization approach from \citet{Ankan2023} to compute the residual
matrices for both $ X $ and $ Y $. For each of the variable $ X $ and $ Y $, we
compute the residual $ R_X $ and $R_Y$ respectively as follows depending on
their type.

@@

We repeat the above steps for both the variables $ X $ and  $ Y $ to get
residuals $ R_X $ and $ R_Y $. Here, we have the option to choose the
estimators based on the properties of the data. Non-parametric ensemble
estimators such as Random Forest and XGBoost can handle any combination of data
types while performing reasonably well. For linear datasets, we can also choose
variants of linear regression depending on the type of data.

@@

Many measures of association based on canonical correlation have been used in
multivariate statistics, each with slightly different properties.
\begin{itemize}
	\item Wilks' Lambda: $\Lambda = \prod_{i}^{\min(p, q)} (1 - \phi_i^2) $
	\item Roy's Largest Root: $ \theta = \max_i(\phi_i^2) $
	\item Pillai's Trace: $ \tau = \sum_{i=1}^{\min(p, q)} \phi_i^2 $
\end{itemize}
